Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2022-03-10 supplement.tex
address some comments
Update supplement.tex
Added section about relating model predictions to morphological variation

\subsection{What does this mean for expectation of morphology btw branches?}
\label{morpho_diff}
\aprcomment{Tim to fill in, and edit section header}

>>>>

\subsection{Relating model predictions to morphological variation} \label{sec:morpho_diff}

Here we show how the predictions of our model can be related to morphological variation in the fossil record. Let $\Delta^2$ be the expected squared difference in a morphological measurement between fossil 1 and fossil 2.

\begin{equation}
	\Delta^2=\mathrm{E}\left\{\left(\bar{X}_1-\bar{X}_2\right)^2\right\}+\mathrm{E}\left\{V_1\right\}+\mathrm{E}\left\{V_2\right\}
\end{equation}

where $\bar{X}_1$ and $V_1$ and $\bar{X}_2$ and $V_2$ are measurement means and variances for the populations from which fossil 1 and fossil 2 are sampled. If we assume that the trait being measured is evolving neutrally, which seems reasonable as an approximation based on studies of morphological (cranial) variation in humans and Neanderthals \tdwcomment{cite here: main text references 36,37, and 38}, then

\begin{equation}
	\mathrm{E}\left\{\left(\bar{X}_1-\bar{X}_2\right)^2\right\}=4\sigma_m^2\tau_B-2\sigma_m^2\tau_{W_1}-2\sigma_m^2\tau_{W_2}
\end{equation}

\begin{equation}
\mathrm{E}\left\{V_A\right\}=\frac{1}{h^2}\sigma_m^2\tau_{W_1}
\end{equation}

\begin{equation}
\mathrm{E}\left\{V_B\right\}=\frac{1}{h^2}\sigma_m^2\tau_{W_2}
\end{equation}

where $\sigma^2_m$ is the additive genetic variance introduced by mutation (per zygote per generation), $h^2$ is the (narrow sense) heritability, $\tau_{W_1}$ and $\tau_{W_2}$ are the average coalescence time of pairs of alleles from the same population and $\tau_B$ is the average coalescence time of pairs of alleles from different populations \tdwcomment{cite here: Whitlock 1999 doi.org/10.1017/S0016672399004127 and Weaver 2016 doi.org/10.1002/ece3.2522}. Substituting and simplifying gives

\begin{equation}
	\Delta^2=\sigma_m^2 \left( 4\tau_B+\frac{1-2h^2}{h^2} \left( \tau_{W_1}+\tau_{W_2} \right) \right)
\end{equation}

which relates $\Delta^2$ to between and within population coalescence times. As a further approximation, if we assume that the heritability of the measurement is close to one half, then

\begin{equation}
	\Delta^2\approx4\sigma_m^2\tau_B
\end{equation}

which shows that under these assumptions $\Delta^2$ is proportional to $\tau_B$.

Nucleotide diversity, $\pi$, for pairs of individuals sampled from different populations is also proportional to $\tau_B$. Therefore, ratios of $\Delta^2$ can be predicted from ratios of $\pi$ for the same comparisons. For example, a ratio for which $\pi$ for stem 1 and stem 2 is the numerator and $\pi$ for a pair of contemporary human populations is the denominator can be used to predict $\Delta^2$ for two fossils sampled from each of the two stems based on $\Delta^2$ for contemporary humans. In this way, the predictions of our model can be related to morphological variation in the fossil record.
